Two applications of logic programming to Coq

### 2020-11-07 papers/types2020/elab.tex
Finished a first pass over Sections 3 and 4.  I give up the lock on section 3.  I wish to continue with Section 4 a bit more.
Moved around large pieces of text as discussed this morning.

\section{Elaboration of external proof certificates for the Coq
  kernel}

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, the formula corresponding to
that type is, in fact, accepted by Coq users as a theorem of
intuitionistic logic along with any axioms that have been adopted.
The rest of the Coq system, especially its tactic language, is
designed to help a human user build proofs-as-$\lambda$-terms that can
be checked by the kernel.

@@

As we have describe in Section~\ref{sec:three}, the Foundational Proof
Certificate (FPC) framework has been designed for formally defining a
rich collection of proof certificates that do not necessarily contain
full details.  Proof checkers for FPC certificates can be written
using simple logic programs: the standard implementation mechanisms of
unification and backtracking search can then be used to reconstruct
missing proof details \cite{miller17fac}.  Such proof checking kernels
can also be used to elaborate such certificates into fully detailed
proofs \cite{blanco17phd,blanco17cade}.

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.
%
Following the general outline that has been described in
\cite{blanco17cade,blanco20coq}, we assume that the following step are
taken.
%
\begin{enumerate}
\item Have an external prover output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

@@

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
%
It is also possible to pair two different proof certificates, defined
by two different FPC definitions, and do the proof checking in
parallel.  The logic programming setting allows such parallel checking
to also serve as checking-and-synthesizing.  That is, during the
checking of one certificate, it is possible to synthesize, for
example, a fully detailed term that is appropriate for handing to the
Coq kernel.

@@

As an example, consider an external theorem prover for propositional
intuitionistic logic which return proof structures that are De Bruijn
expressions for the underlying untype $\lambda$-proof that encodes
(part of) the proofs it finds.  Consider the De Buijn expression
example at the end of Section~\ref{ssec:beyond}.  This can be used as
evidence that the following is a theorem of Coq.

@@

Of course, the proof information in the untyped term must be
elaborated to be a fully typed and functional expression.  Proof
checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
of) the proof Figure~\ref{fig:kernel}.

\begin{metanote}
  AM: I think we need more: the FPC may go to the appendix, we may want to say that there is a tactic that proves these theorems etc. 
\end{metanote}

%\newpage

@@

% LocalWords:  intuitionistic

>>>>

\section{Elaboration of external proof certificates for the Coq kernel}
\label{sec:elab}

The trusted base of Coq is its kernel, which is a type-checking
program that certifies that a dependently typed $\lambda$-term has a
given type.  If type checking succeeds, the formula corresponding to
that type is, in fact, accepted by Coq users as a theorem of
intuitionistic logic (along with any axioms that have been asserted).
The rest of the Coq system, especially its tactic language, is
designed to help a human user build proofs-as-$\lambda$-terms that can
be checked by the kernel.

@@

As we have describe in Section~\ref{sec:three}, the Foundational Proof
Certificate (FPC) framework has been designed for formally defining a
broad collection of proof certificates that do not necessarily contain
full details.  Proof checkers for FPC certificates can be written
using simple logic programs: the standard implementation mechanisms of
unification and backtracking search can then be used to reconstruct
missing proof details \cite{miller17fac}.  Such proof checking kernels
can also be used to elaborate such certificates into fully detailed
proofs \cite{blanco17phd,blanco17cade}.

\subsection{Moving beyond Horn clauses}
\label{ssec:beyond}

If we do not restrict our attention to only Horn clauses then there is
no problem presenting a focused proof system for all of first-order
intuitionistic logic using the connectives $\forall$ and $\supset$.
The following rules extend those given earlier in
Section~\ref{ssec:focused}. 
\[
  \infer[\supset R]{\Pscr\vdash B\supset D}{\Pscr, B\vdash D}
  \qquad
  \infer[\forall R]{\Pscr\vdash \forall x.D}{\Pscr\vdash D[y/x]}
\]
As usual, the $\forall R$ rule has the restriction that $y$ is not
free in its conclusion.

As has been detailed in earlier work on foundational proof
certificates, this richer notion of proof system can provide for
richer proof certificates.  The main differences with what we have
seen before is that the left-hand sides of sequents can now grow
during the proof checking process.  As a result, an indexing mechanism
is used so that it is possible for the decide expert to properly
select locally added assumptions instead of just those assumptions
accepted globally.

To illustrate a proof certificate format that uses this richer logic
and proof system, consider the simple \lP code in
Figure~\ref{fig:debruijn}.  Using the constants provided in that
figure, the untyped $\lambda$-term
$\lambda x (x (\lambda y (y (\lambda z (x (\lambda u\,z))))))$ can be
encoded as the following Elpi term of type \lsti{deb}.
%
\begin{lstlisting}[basicstyle=\ttfamily,language=lprolog]
(lambda (apply 0 [lambda (apply 0 [lambda (apply 2 
                                                    [lambda (apply 1 [])])])]))
\end{lstlisting}
Using the constructor \lsti{lc} and \lsti{args}, terms in De Bruijn
syntax (terms of type \lsti{deb}) are incorporated into proof
certificates (terms of type \lsti{cert}) along with other integer
arguments that are needed to compute offsets to address bound
variables.  
%
This FPC can be interpreted by the proof certificate checker in the
file \verb+ljf-dep.mod+ of the repository
\url{https://github.com/proofcert/fpc-elpi/}.  We note that this proof
certificate checker is also able to simultaneously synthesize a Coq
proof term while checking (see Section~\ref{sec:elab}) all the while
remaining less than 90 lines of Elpi code.

\begin{figure}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.sig}
\lstinputlisting[basicstyle=\ttfamily,language=lprolog,linerange={start-end}]{code/deb-fpc.mod}
\caption{The FPC definition of De Bruijn notation as proof evidence.}
\label{fig:debruijn}
\end{figure}

****************

We describe here briefly how to use the technology behind FPCs and
logic programming in order to provide a flexible approach to
connecting external provers of first-order intuitionistic logic to
Coq.
%
Following the general outline that has been described in
\cite{blanco17cade,blanco20coq}, we assume that the following step are
taken.
%
\begin{enumerate}
\item Modify an external prover to output some form of proof evidence
  (proof certificate) for formulas it claims are theorems.

@@

\item Check proofs by executing the logic programming checker that is
  parameterized by the particular FPC definition.
\end{enumerate}
%
It is also possible to pair together two different proof certificates,
defined by two different FPC definitions, and do the proof checking in
parallel.  The logic programming setting allows such parallel checking
to also serve as checking-and-synthesizing.  That is, during the
checking of one certificate, it is possible to synthesize, for
example, a fully detailed term that is appropriate for handing to the
Coq kernel.

@@

As an example, consider an external theorem prover for propositional
intuitionistic logic which returns proof structures as untyped
$\lambda$-terms using De Bruijn's notation \cite{}.  For example, the
De Bruijn expression at the end of Section~\ref{ssec:beyond} can be
used as evidence that the following is a theorem of Coq.

@@

Of course, the proof information in the untyped term must be
elaborated to be a fully typed and functional expression.  Proof
checking using the FPC in Figure~\ref{fig:debruijn} and (an extension
of) the proof checker in Figure~\ref{fig:kernel}.

@@

% LocalWords:  intuitionistic dneg peirce forall
